\pkg{casebase}: An Alternative Framework For Survival Analysis and Comparison of Event Rates

### 2020-08-05 analysis/paper/paper.Rmd
update cs 3&4

---
documentclass: jss
author:
  - name: Sahir Rai Bhatnagar*
    affiliation: McGill University
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{sahir.bhatnagar@mail.mcgill.ca}
    url: http://sahirbhatnagar.com/
  - name: Maxime Turgeon*
    affiliation: 'University of Manitoba \AND'
    address: >
      186 Dysart Road
      Winnipeg, MB, Canada R3T 2N2
    email: \email{max.turgeon@umanitoba.ca}
    url: https://maxturgeon.ca/
  - name: Jesse Islam 
    affiliation: McGill University
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{jesse.islam@mail.mcgill.ca}
  - name: James Hanley
    affiliation: McGill University
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{james.hanley@mcgill.ca}
    url: http://www.medicine.mcgill.ca/epidemiology/hanley/
  - name: Olli Saarela
    affiliation: University of Toronto
    address: >
      Dalla Lana School of Public Health, 
      155 College Street, 6th floor, 
      Toronto, Ontario 
      M5T 3M7, Canada
    email: \email{olli.saarela@utoronto.ca}
    url: http://individual.utoronto.ca/osaarela/
title:
  formatted: "\\pkg{casebase}: An Alternative Framework For Survival Analysis"
  # If you use tex in the formatted title, also supply version without
  plain:     "casebase: An Alternative Framework For Survival Analysis"
  # For running headers, if needed
  short:     "\\pkg{casebase}: An Alternative Framework For Survival Analysis"
abstract: >
  The abstract of the article. *SRB and MT contributed equally to this work.
keywords:
  # at least one keyword must be supplied
  formatted: [survival analysis, absolute risk, data visualization]
  plain:     [survival analysis, absolute risk, data visualization]
preamble: >
  \usepackage{amsmath}
  \usepackage{longtable}
  \usepackage{graphicx}
  \usepackage{tabularx}
  \usepackage{float}
  \usepackage{booktabs}
  \usepackage{makecell}
  \usepackage{tabu}
  \DeclareUnicodeCharacter{2500}{-}
output: 
  rticles::jss_article:
    number_sections: TRUE     #added argument option 
    citation_package: "natbib"  #All my citations use biblatex, not natbib. 
biblio-style: jss      #Listed to use in JSS Instructions for Authors, but not in template by default. 
bibliography: references.bib  #Also not included in template by default. 
fig_width: 12
fig_height: 9
fig_caption: true
editor_options: 
  chunk_output_type: console
---

@@

eval_introduction <- TRUE
eval_theory <- TRUE
eval_implementation <- TRUE
eval_cs1 <- T
eval_cs2 <- T
eval_cs3 <- TRUE
eval_cs4 <- T
eval_colophon <- TRUE

@@

Survival analysis has been greatly influenced over the last 50 years by the partial likelihood approach of the Cox proportional hazard model [@cox1972regression]. This approach provides a flexible way of assessing the influence of covariates on the hazard function, without the need to specify a parametric survival model. This flexibility comes at the cost of decoupling the baseline hazard from the effect of the covariates. To recover the whole survival curve---or the cumulative incidence function (CIF)---we then need to separately estimate the baseline hazard [@breslow1972discussion]. This in turn often leads to stepwise estimates of the survival function.

@@

With the goal of fitting smooth-in-time hazard functions, Hanley & Miettinen [-@hanley2009fitting] proposed a general framework for estimating fully parametric hazard models via logistic regression. Their approach provides users that are familiar with generalized linear models with a natural way of fitting parametric survival models. Moreover, their framework is very flexible: general functions of time can be estimated (e.g. using splines or general additive models), and hence these models retain some of the flexibility of partial likelihood approaches.

In this article, we present an \proglang{R} package that combines the ideas of Hanley & Miettinen into a simple interface. The purpose of the \pkg{casebase} package is to provide practitioners with an easy-to-use software tool to predict the risk (or cumulative incidence) of an event, conditional on a particular patient's covariate profile. Our package retains the flexibility of case-base sampling and the familiar interface of the \code{glm} function. In addition, we provide extensive visualization tools.

@@

As discussed in Hanley & Miettinen [-@hanley2009fitting], the key idea behind case-base sampling is to discretize the study base into an infinite amount of *person moments*. These person moments are indexed by both an individual in the study and a time point, and therefore each person moment has a covariate profile, an exposure status and an outcome status attached to it. We note that there is only a finite number of person moments associated with the event of interest (what Hanley & Miettinen call the *case series*). Case-base sampling refers to the sampling from the base of a representative finite sample called the *base series*.

To describe the theoretical foundations of case-base sampling, we use the framework of counting processes. In what follows, we abuse notation slightly and omit any mention of $\sigma$-algebras; the interested reader can refer to Saarela & Arjas [-@saarela2015non] and Saarela [-@saarela2016case]. First, let $N_{i}(t) \in \{0, 1\}$ be counting processes corresponding to the event of interest for individual $i=1, \ldots,n$. For simplicity, we will consider Type I censoring due to the end of follow-up at time $\tau$ (the general case of non-informative censoring is treated in Saarela [-@saarela2016case]). We assume a continuous time model, which implies that the counting process jumps are less than or equal to one. We are interested in modeling the hazard functions $\lambda_{i}(t)$ of the processes $N_i(t)$, and which satisfy
$$\lambda_{i}(t) dt = E[dN_{i}(t)].$$

Next, we model the base series sampling mechanism using non-homogeneous Poisson processes $R_i(t) \in \{0, 1, 2, \ldots\}$, with the person-moments where $dR_i(t) = 1$ constituting the base series. The process $Q_{i}(t) = R_i(t) + N_{i}(t)$ then counts both the case and base series person-moments contributed by individual $i$. This process is typically defined by the user via its hazard function $\rho_i(t)$. The process $Q_{i}(t)$ is characterized by $E[dQ_{i}(t)] = \lambda_{i}(t)dt + \rho_i(t)dt$.

If the hazard function $\lambda_{i}(t; \theta)$ is parametrized in terms of $\theta$, we could define an estimator $\hat{\theta}$ by maximization of the likelihood expression
$$L_0(\theta) = \prod_{i=1}^n \exp\left\{ -\int_0^\tau \lambda_i(t; \theta) dt \right\} \prod_{i=1}^{n} \prod_{t\in[0,\tau)} \lambda_{i}(t;\theta)^{dN_{i}(t)}\left(1 - \lambda_{i}(t;\theta)\right)^{1 - dN_{i}(t)},$$
where $\prod_{t\in[0,u)}$ represents a product integral from $0$ to $u$. However, the integral over time makes the computation and maximization of $L_0(\theta)$ challenging.

Case-base sampling allows us to avoid this integral. By conditioning on a sampled person-moment, we get individual contributions of the form
$$P(dN_{i}(t) \mid dQ_{i}(t) = 1) \stackrel{\theta}{\propto} \frac{\lambda_{i}(t; \theta)^{dN_{i}(t)}}{\rho_i(t) + \lambda_{i}(t;\theta)}.$$
Therefore, we can define an estimating equation for $\theta$ as follows:
$$L(\theta) = \prod_{i=1}^{n} \prod_{t\in[0,\tau)} \left(\frac{\lambda_{i}(t; \theta)^{dN_{i}(t)}}{\rho_i(t) + \lambda_{i}(t;\theta)}\right)^{dQ_i(t)}.$$
When a logarithmic link function is used for modeling the hazard function, the above expression is of a logistic regression form with an offset term $\log(1/\rho_i(t))$. Note that the sampling units selected in the case-base sampling mechanism are person-moments, rather than individuals, and the parameters to be estimated are hazards or hazard ratios rather than odds or odds ratios. Generally, an individual can contribute more than one person-moment, and thus the terms in the product integral are not independent. Nonetheless, Saarela [-@saarela2016case] showed that the logarithm of this estimating function has mean zero at the true value $\theta=\theta_0$, and that the resulting estimator $\hat{\theta}$ is asymptotically normally distributed.

@@

Case-base sampling can also be used in the context of competing-risk analyses. Assuming there are $J$ competing events, we can show that each person-moment's contribution of the likelihood is of the form

@@

where $N_j(t)$ is the counting process associated with the event of type $j$ and $\lambda_j(t)$ is the corresponding hazard function. As may be expected, this functional form is similar to the terms appearing in the likelihood function for multinomial regression.\footnote{Specifically, it corresponds to the following parametrization: \begin{align*} \log\left(\frac{P(Y=j \mid X)}{P(Y = J \mid X)}\right) = X^T\beta_j, \qquad j = 1,\ldots, J-1.\end{align*}}

@@

Parametric survival models are implemented in a handful of packages: \pkg{CFC} [-@mahani2015bayesian], \pkg{flexsurv} [-@flexsurv], \pkg{SmoothHazard} [-@smoothHazard], \pkg{rsptm2} [-@clements_liu], \pkg{mets} [-@scheike2014estimating], and \pkg{survival} [-@survival-package]. The types of models they allow vary for each package. For example, \pkg{SmoothHazard} is limited to Weibull distributions [-@smoothHazard], whereas both \pkg{flexsurv} and \pkg{survival} allow users to supply any distribution of their choice. Also, \pkg{flexsurv}, \pkg{smoothhazard}, \pkg{mets} and \pkg{rstpm2} also have the ability to model the effect of time using splines, which allows flexible modeling of the hazard function. Moreover, \pkg{flexsurv} has the ability to estimate both scale and shape parameters for a variety of parametric families. As discussed above, \pkg{casebase} can model any parametric family whose log-hazard can be expressed as a linear combination of covariates (including time). Therefore, our package allows the user to model the effect of time using splines. Also, by including interaction terms between covariates and time, it also allows users to fit time-varying coefficient models. However, we do not explicitly model any shape parameter, unlike \pkg{flexsurv}.

Several packages implement penalized estimation for the Cox model: \pkg{glmnet} [-@regpathcox], \pkg{glmpath} [-@park_hastie], \pkg{penalized} [-@l1penal], \pkg{RiskRegression} [-@gerds_blanche]. Moreover, some packages also include penalized estimation in the context of Cox models with time-varying coefficients: elastic-net penalization with \pkg{CoxRidge} [-@perperoglou] and \pkg{rstpm2} [-@clements_liu], while \pkg{survival} [-@survival-package] has an implementation of ridge-penalized estimation. On the other hand, our package \pkg{casebase} provides penalized estimation in the context of parametric hazards. To our knowledge, \pkg{casebase} and \pkg{rsptm2} are the only packages to offer this functionality.

Next, several \proglang{R} packages implement methodologies for competing risk analysis; for a different perspective on this topic, see Mahani & Sharabiani [-@mahani2015bayesian]. The package \pkg{cmprsk} provides methods for cause-specific subdistributions, such as in the Fine-Gray model [-@fine1999proportional]. On the other hand, the package \pkg{CFC} estimates cause-specific CIFs from unadjusted, non-parametric survival functions. Our package \pkg{casebase} also provides functionalities for competing risk analysis by estimating parametrically the cause-specific hazards. From these quantities, we can then estimate the cause-specific CIFs.

Finally, several packages include functions to estimate the CIF. The corresponding methods generally fall into two categories: transformation of the estimated hazard function, and semi-parametric estimation of the baseline hazard. The first category broadly corresponds to parametric survival models, where the full hazard is explicitly modeled. Using this estimate, the survival function and the CIF can be obtained using their functional relationships (see Equations \ref{eqn:surv} and \ref{eqn:CI} below). Packages providing this functionality include \pkg{CFC}, \pkg{Flexsurv}, \pkg{mets}, and \pkg{survival}. Our package \pkg{casebase} also follows this approach for both single-event and competing-risk analyses. The second category outlined above broadly corresponds to semi-parametric models. These models do not model the full hazard function, and therefore the baseline hazard needs to be estimated separately in order to estimate the survival function. This is achieved using semi-parametric estimators (e.g. Breslow's estimator) or parametric estimators (e.g. spline functions). Packages that implement this approach include \pkg{RiskRegression}, \pkg{rstpm2}, and \pkg{survival}. As mentioned in the introduction, a key distinguishing factor between these two approaches is that the first category leads to smooth estimates of the cumulative incidence function, whereas the second category often produces estimates in the form of step-wise functions. Providing smooth estimates of the CIF was one of the main motivations for introducing case-base sampling in survival analysis.

@@

\textbf{Package}        & \textbf{Risks}  & \textbf{Non PH} & \textbf{Regression} & \textbf{Splines} & \textbf{Parametric} & \textbf{Parametric} & \textbf{Censoring} & \textbf{Risk}           \\

@@

\textbf{crrp}            & x                        &                           & x                     &                  &                     & x                        &                                  &                                  \\

@@

\textbf{RiskRegression}  & x                         &                           & x                     &                  &                     & x                        &                                  & x                                \\
\textbf{Rstpm2}          &                          & x                         &                      & x                & x                   & x                        & x                                & x                         \\

@@

\caption{Comparison of various \proglang{R} packages for survival analysis based on several defining features. Competing Risks: whether or not an implementation for competing risks is present. Allows Non-PH: permits models that can incorporate variables that change with survival time. Penalized Regression: allows for a penalty term when fitting the model. This is not restricted to lasso nor ridge. Splines: the package permits a flexible fit on time through the use of splines. Parametric: the package has implementations for parametric models. Semi-parametric: the package has implementations for semi-parametric models. Interval/left censoring: the package can interpret interval and left-censoring. If this is not selected, the package only handles right-censoring. Absolute Risk: the package has an implementation for survival curves, cumulative incidence or cumulative hazard rate functions to be plotted.}

@@

Figure \ref{fig:plot-stratified-erspc-data} is built sequentially by first adding a layer for the area representing the population time in gray, with subjects having the least amount of observation time plotted at the top of the y-axis. We immediately notice a distinctive *step-wise shape* in the population time area. This is due to the randomization of the Finnish cohorts which were carried out on January 1 of each of year from 1996 to 1999. Coupled with the uniform December 31 2006 censoring date, this led to large numbers of men with exactly 11, 10, 9 or 8 years of follow-up. Tracked backwards in time (i.e. from right to left), the population-time plot shows the recruitment pattern from its beginning in 1991, and the January 1 entries in successive years. Tracked forwards in time (i.e. from left to right), the plot for the first three years shows attrition due entirely to death (mainly from other causes). Since the Swedish and Belgian centres were the last to complete recruitment--in December 2003--the minimum potential follow-up is three years. Tracked further forwards in time (i.e. after year 3) the attrition is a combination of deaths and staggered entries. As we can see, population-time plots summarise a wealth of information about the study into a simple graph.

@@

In the case of multiple events, the hazard is fitted via multinomial regression as performed by the \pkg{VGAM} package. We selected this package for its ability to fit multinomial regression models with an offset.

@@

$$ S(t) = \exp\left(-\int_0^t \Lambda(u;X) du\right),\qquad \Lambda(t;X) = \sum_{j=1}^J \lambda_j(t;X).$$
From this, we can derive the event-specific subdensities:

@@

The data contains information on 177 patients who received a stem-cell transplant for acute leukemia. The event of interest is relapse, but other competing causes (e.g. transplant-related death) were also recorded. Several covariates were captured at baseline: sex, disease type (acute lymphoblastic or myeloblastic leukemia, abbreviated as ALL and AML, respectively), disease phase at transplant (Relapse, CR1, CR2, CR3), source of stem cells (bone marrow and peripheral blood, coded as BM+PB, or only peripheral blood, coded as PB), and age. <!--A summary of these baseline characteristics appear in Table \ref{tab:table1bmtcrr}. We note that the statistical summaries were generated differently for different variable types: for continuous variables, we gave the range, followed by the mean and standard deviation; for categorical variables, we gave the counts for each category.-->

@@

Our main objective is to compute the absolute risk of relapse for a given set of covariates. We start by fitting a smooth hazard to the data using a linear term for time:

@@

We will compare our estimates to that obtained from a corresponding Fine-Gray model [-@fine1999proportional]. The Fine-Gray model is a semiparametric model for the cause-specific *subdistribution*, i.e. the function $f_k(t)$ such that
$$CI_k(t) =1 - \exp\left( - \int_0^t f_k(u) \textrm{d}u \right),$$
where $CI_k(t)$ is the cause-specific cumulative incidence. The Fine-Gray model allows to directly assess the effect of a covariate on the subdistribution, as opposed to the hazard. For the computation, we will use the \pkg{timereg} package [@timereg]:

@@

# 1. Cox with only sps
coxSPS <- survival::coxph(Surv(time = d.time, event = death) ~ sps, 
                          data = train, x = TRUE)
abcoxSPS <- survival::survfit(coxSPS, newdata = test)
# 2. Cox with only aps
coxAPS <- survival::coxph(Surv(time = d.time, event = death) ~ aps, 
                          data = train, x = TRUE)
abcoxAPS <- survival::survfit(coxAPS, newdata = test)
# 3. Cox with everything but sps and aps
coxFull <- survival::coxph(Surv(time = d.time, event = death) ~ . - aps - sps, 
                           data = train, x = TRUE)
abcoxFull <- survival::survfit(coxFull, newdata = test)
# 4. Kaplan-Meier
KM <- survival::coxph(Surv(time = d.time, event = death) ~ 1, data = test,x=TRUE)
abKM <- survival::survfit(KM)
abKMcompare <- data.frame(
  Time = abKM$time,
  KM = 1 - abKM$surv
)

@@

```{r baselines, echo=FALSE, eval=eval_cs3}
cox_ci <- ggplot(baseCoxComparisonData, aes(Time)) +
  geom_step(aes(y = APS, colour = "APS")) +
  geom_step(aes(y = Full, col = "Full")) +
  geom_step(aes(y = SPS, colour = "SPS")) +
  geom_step(data = abKMcompare, mapping = aes(x = Time, y = KM, color = "K-M")) +
  labs(y = "Probability of Death", x = "Follow-up time (Years)", color = "Models") +
  scale_color_manual(values = q4, breaks = c("APS", "Full", "SPS", "K-M")) + 
  paper_gg_theme
```

>>>>

---
documentclass: jss
author:
  - name: Sahir Rai Bhatnagar*
    affiliation: McGill University
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{sahir.bhatnagar@mail.mcgill.ca}
    url: http://sahirbhatnagar.com/
  - name: Maxime Turgeon*
    affiliation: 'University of Manitoba \AND'
    address: >
      186 Dysart Road
      Winnipeg, MB, Canada R3T 2N2
    email: \email{max.turgeon@umanitoba.ca}
    url: https://maxturgeon.ca/
  - name: Jesse Islam 
    affiliation: McGill University
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{jesse.islam@mail.mcgill.ca}
  - name: James A. Hanley
    affiliation: McGill University
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{james.hanley@mcgill.ca}
    url: http://www.medicine.mcgill.ca/epidemiology/hanley/
  - name: Olli Saarela
    affiliation: University of Toronto
    address: >
      Dalla Lana School of Public Health, 
      155 College Street, 6th floor, 
      Toronto, Ontario 
      M5T 3M7, Canada
    email: \email{olli.saarela@utoronto.ca}
    url: http://individual.utoronto.ca/osaarela/
title:
  formatted: "\\pkg{casebase}: An Alternative Framework For Survival Analysis and Comparison of Event Rates"
  # If you use tex in the formatted title, also supply version without
  plain:     "casebase: An Alternative Framework For Survival Analysis and Comparison of Event Rates"
  # For running headers, if needed
  short:     "\\pkg{casebase}: An Alternative Framework For Survival Analysis"
abstract: >
  The abstract of the article. *SRB and MT contributed equally to this work.
keywords:
  # at least one keyword must be supplied
  formatted: [survival analysis, absolute risk, data visualization]
  plain:     [survival analysis, absolute risk, data visualization]
preamble: >
  \usepackage{amsmath}
  \usepackage{longtable}
  \usepackage{graphicx}
  \usepackage{tabularx}
  \usepackage{float}
  \usepackage{booktabs}
  \usepackage{makecell}
  \usepackage{tabu}
  \DeclareUnicodeCharacter{2500}{-}
output: 
  rticles::jss_article:
    number_sections: TRUE     #added argument option 
    citation_package: "natbib"  #All my citations use biblatex, not natbib. 
biblio-style: jss      #Listed to use in JSS Instructions for Authors, but not in template by default. 
bibliography: references.bib  #Also not included in template by default. 
fig_width: 12
fig_height: 9
fig_caption: true
editor_options: 
  chunk_output_type: console
---

@@

eval_introduction <- TRUE
eval_theory <- TRUE
eval_implementation <- TRUE
eval_cs1 <- TRUE
eval_cs2 <- TRUE
eval_cs3 <- TRUE
eval_cs4 <- TRUE
eval_colophon <- TRUE

@@

Survival analysis and the comparison of event rates has been greatly influenced over the last 50 years by the partial likelihood approach of the Cox proportional hazard model [@cox1972regression]. This approach provides a flexible way of assessing the influence of covariates on the hazard function, without the need to specify a parametric survival model. This flexibility comes at the cost of decoupling the baseline hazard from the effect of the covariates. To recover the whole survival curve---or the cumulative incidence function (CIF)---we then need to separately estimate the baseline hazard [@breslow1972discussion]. This in turn leads to stepwise estimates of the survival function.

@@

With the goal of fitting smooth-in-time hazard functions, Hanley & Miettinen [-@hanley2009fitting] proposed a general framework for estimating fully parametric hazard models via logistic regression. Their approach provides users that are familiar with generalized linear models with a natural way of fitting parametric survival models. Moreover, their framework is very flexible: general functions of time can be estimated (e.g. using splines or general additive models), and hence these models retain some of the flexibility of Cox's partial likelihood approach.

In this article, we present an \proglang{R} package that combines the ideas of Hanley & Miettinen into a simple interface. The purpose of the \pkg{casebase} package is to provide practitioners with an easy-to-use software tool to compute a patient's risk (or cumulative incidence) of an event, conditional on a particular patient's covariate profile. Our package retains the flexibility of case-base sampling and the familiar interface of the \code{glm} function. In addition, we provide extensive visualization tools.

@@

As discussed in Hanley & Miettinen [-@hanley2009fitting], the key idea behind case-base sampling is to sample from the study base a finite amount of *person moments*. These person moments are indexed by both an individual in the study and a time point, and therefore each person moment has a covariate profile, an exposure status and an outcome status attached to it. We note that there is only a finite number of person moments associated with the event of interest (what Hanley & Miettinen call the *case series*). Case-base sampling refers to the sampling from the base of a representative finite sample called the *base series*.

To describe the theoretical foundations of case-base sampling, we use the framework of counting processes. In what follows, we abuse notation slightly and omit any mention of $\sigma$-algebras. Instead, following Aalen \emph{et al} [-@aalen2008survival], we use the placeholder $\mathrm{past}$ to denote the past history of the corresponding process. The reader interested in more details can refer to Saarela & Arjas [-@saarela2015non] and Saarela [-@saarela2016case]. First, let $N_{i}(t) \in \{0, 1\}$ be counting processes corresponding to the event of interest for individual $i=1, \ldots,n$. For simplicity, we will consider Type I censoring due to the end of follow-up at time $\tau$ (the general case of non-informative censoring is treated in Saarela [-@saarela2016case]). We assume a continuous time model, which implies that the counting process jumps are less than or equal to one. We are interested in modeling the hazard functions $\lambda_{i}(t)$ of the processes $N_i(t)$, and which satisfy
$$\lambda_{i}(t) dt = E[dN_{i}(t)\mid \mathrm{past}].$$

Next, we model the base series sampling mechanism using non-homogeneous Poisson processes $R_i(t) \in \{0, 1, 2, \ldots\}$, with the person-moments where $dR_i(t) = 1$ constituting the base series. The process $Q_{i}(t) = R_i(t) + N_{i}(t)$ then counts both the case and base series person-moments contributed by individual $i$. This process is typically defined by the user via its intensity function $\rho_i(t)$. The process $Q_{i}(t)$ is characterized by $E[dQ_{i}(t)\mid\mathrm{past}] = \lambda_{i}(t)dt + \rho_i(t)dt$.

If the hazard function $\lambda_{i}(t; \theta)$ is parametrized in terms of $\theta$, we could define an estimator $\hat{\theta}$ by maximization of the likelihood expression
$$L_0(\theta) = \prod_{i=1}^n \exp\left\{ -\int_0^{\min(t_i,\tau)} \lambda_i(t; \theta) dt \right\} \prod_{i=1}^{n} \prod_{t\in[0,\tau)} \lambda_{i}(t;\theta)^{dN_{i}(t)},$$
where $\prod_{t\in[0,u)}$ represents a product integral from $0$ to $u$, and where $t_i$ is the event time for individual $i$. However, the integral over time makes the computation and maximization of $L_0(\theta)$ challenging.

Case-base sampling allows us to avoid this integral. By conditioning on a sampled person-moment, we get individual likelihood contributions of the form
$$P(dN_{i}(t) \mid dQ_{i}(t) = 1,\mathrm{past}) \stackrel{\theta}{\propto} \frac{\lambda_{i}(t; \theta)^{dN_{i}(t)}}{\rho_i(t) + \lambda_{i}(t;\theta)}.$$
Therefore, we can define an estimating function for $\theta$ as follows:
$$L(\theta) = \prod_{i=1}^{n} \prod_{t\in[0,\tau)} \left(\frac{\lambda_{i}(t; \theta)^{dN_{i}(t)}}{\rho_i(t) + \lambda_{i}(t;\theta)}\right)^{dQ_i(t)}.$$
When a logarithmic link function is used for modeling the hazard function, the above expression is of a logistic regression form with an offset term $\log(1/\rho_i(t))$. Note that the sampling units selected in the case-base sampling mechanism are person-moments, rather than individuals, and the parameters to be estimated are hazards or hazard ratios rather than odds or odds ratios. Generally, an individual can contribute more than one person-moment, and thus the terms in the product integral are not independent. Nonetheless, Saarela [-@saarela2016case] showed that the logarithm of this estimating function has mean zero at the true value $\theta=\theta_0$, and that the resulting estimator $\hat{\theta}$ is asymptotically normally distributed.

@@

Case-base sampling can also be used in the context of competing-risk analyses. Assuming there are $J$ competing events, we can show that each person-moment's contribution to the likelihood is of the form

@@

where $N_j(t)$ is the counting process associated with the event of type $j$ and $\lambda_j(t)$ is the corresponding cause-specific hazard function. As may be expected, this functional form is similar to the terms appearing in the likelihood function for multinomial regression.\footnote{Specifically, it corresponds to the following parametrization: \begin{align*} \log\left(\frac{P(Y=j \mid X)}{P(Y = J \mid X)}\right) = X^T\beta_j, \qquad j = 1,\ldots, J-1.\end{align*}}

@@

Parametric survival models are implemented in a handful of packages: \pkg{CFC} [-@mahani2015bayesian], \pkg{flexsurv} [-@flexsurv], \pkg{SmoothHazard} [-@smoothHazard], \pkg{rsptm2} [-@clements_liu], \pkg{mets} [-@scheike2014estimating], and \pkg{survival} [-@survival-package]. The types of models they allow vary for each package. For example, \pkg{SmoothHazard} is limited to Weibull distributions [-@smoothHazard], whereas both \pkg{flexsurv} and \pkg{survival} allow users to supply any distribution of their choice. Also, \pkg{flexsurv}, \pkg{smoothhazard}, \pkg{mets} and \pkg{rstpm2} also have the ability to model the effect of time using splines, which allows flexible modeling of the hazard function. Moreover, \pkg{flexsurv} has the ability to estimate both scale and shape parameters for a variety of parametric families. As discussed above, \pkg{casebase} can model any parametric family whose log-hazard can be expressed as a linear combination of covariates (including time). Therefore, our package allows the user to model the effect of time using splines. Also, by including interaction terms between covariates and time, it also allows users to fit (non-proportional) time-varying coefficient models. However, we do not explicitly model any shape parameter, unlike \pkg{flexsurv}.

Several packages implement penalized estimation for the Cox model: \pkg{glmnet} [-@regpathcox], \pkg{glmpath} [-@park_hastie], \pkg{penalized} [-@l1penal], \pkg{RiskRegression} [-@gerds_blanche]. Moreover, some packages also include penalized estimation in the context of Cox models with time-varying coefficients: elastic-net penalization with \pkg{CoxRidge} [-@perperoglou] and \pkg{rstpm2} [-@clements_liu], while \pkg{survival} [-@survival-package] has an implementation of ridge-penalized estimation. On the other hand, our package \pkg{casebase} provides penalized estimation of the hazard function. To our knowledge, \pkg{casebase} and \pkg{rsptm2} are the only packages to offer this functionality.

Next, several \proglang{R} packages implement methodologies for competing risk analysis; for a different perspective on this topic, see Mahani & Sharabiani [-@mahani2015bayesian]. The package \pkg{cmprsk} provides methods for cause-specific subdistribution hazards, such as in the Fine-Gray model [-@fine1999proportional]. On the other hand, the package \pkg{CFC} estimates cause-specific CIFs from unadjusted, non-parametric survival functions. Our package \pkg{casebase} also provides functionalities for competing risk analysis by estimating parametrically the cause-specific hazards. From these quantities, we can then estimate the cause-specific CIFs.

Finally, several packages include functions to estimate the CIF. The corresponding methods generally fall into two categories: transformation of the estimated hazard function, and semi-parametric estimation of the baseline hazard. The first category broadly corresponds to parametric survival models, where the full hazard is explicitly modeled. Using this estimate, the survival function and the CIF can be obtained using their functional relationships (see Equations \ref{eqn:surv} and \ref{eqn:CI} below). Packages providing this functionality include \pkg{CFC}, \pkg{Flexsurv}, \pkg{mets}, and \pkg{survival}. Our package \pkg{casebase} also follows this approach for both single-event and competing-risk analyses. The second category outlined above broadly corresponds to semi-parametric models. These models do not model the full hazard function, and therefore the baseline hazard needs to be estimated separately in order to estimate the survival function. This is achieved using semi-parametric estimators (e.g. Breslow's estimator) or parametric estimators (e.g. spline functions). Packages that implement this approach include \pkg{RiskRegression}, \pkg{rstpm2}, and \pkg{survival}. As mentioned in the introduction, a key distinguishing factor between these two approaches is that the first category leads to smooth estimates of the cumulative incidence function, whereas the second category often produces estimates in the form of stepwise functions. Providing smooth estimates of the CIF was one of the main motivations for introducing case-base sampling in survival analysis.

@@

\textbf{Package}        & \textbf{Risks}  & \textbf{Non PH} & \textbf{Likelihood} & \textbf{Splines} & \textbf{Parametric} & \textbf{Parametric} & \textbf{Censoring} & \textbf{Risk}           \\

@@

\textbf{crrp}            & x                        &                           & x                     &                  &                     &                          &                                  &                                  \\

@@

\textbf{RiskRegression}  &                          &                           & x                     &                  &                     & x                        &                                  & x                                \\
\textbf{Rstpm2}          &                          & x                         & x                     & x                & x                   & x                        & x                                & x                         \\

@@

\caption{Comparison of various \proglang{R} packages for survival analysis based on several defining features.}

@@

Figure \ref{fig:plot-stratified-erspc-data} is built sequentially by first adding a layer for the area representing the population time in gray, with subjects having the least amount of observation time plotted at the top of the y-axis. We immediately notice a distinctive *stepwise shape* in the population time area. This is due to the randomization of the Finnish cohorts which were carried out on January 1 of each of year from 1996 to 1999. Coupled with the uniform December 31 2006 censoring date, this led to large numbers of men with exactly 11, 10, 9 or 8 years of follow-up. Tracked backwards in time (i.e. from right to left), the population-time plot shows the recruitment pattern from its beginning in 1991, and the January 1 entries in successive years. Tracked forwards in time (i.e. from left to right), the plot for the first three years shows attrition due entirely to death (mainly from other causes). Since the Swedish and Belgian centres were the last to complete recruitment--in December 2003--the minimum potential follow-up is three years. Tracked further forwards in time (i.e. after year 3) the attrition is a combination of deaths and staggered entries. As we can see, population-time plots summarise a wealth of information about the study into a simple graph.

@@

In the case of multiple competing events, the hazard is fitted via multinomial regression as performed by the \pkg{VGAM} package. We selected this package for its ability to fit multinomial regression models with an offset.

@@

$$ S(t) = \exp\left(-\int_0^t \lambda(u;X) du\right),\qquad \lambda(t;X) = \sum_{j=1}^J \lambda_j(t;X).$$
From this, we can derive the event-specific subdensities:

@@

The data contains information on 177 patients who received a stem-cell transplant for acute leukemia. The event of interest is relapse, but other competing causes (e.g. death, progression, graft failure, graft-versus-host disease) were also recorded. Several covariates were captured at baseline: sex, disease type (acute lymphoblastic or myeloblastic leukemia, abbreviated as ALL and AML, respectively), disease phase at transplant (Relapse, CR1, CR2, CR3), source of stem cells (bone marrow and peripheral blood, coded as BM+PB, or only peripheral blood, coded as PB), and age. <!--A summary of these baseline characteristics appear in Table \ref{tab:table1bmtcrr}. We note that the statistical summaries were generated differently for different variable types: for continuous variables, we gave the range, followed by the mean and standard deviation; for categorical variables, we gave the counts for each category.-->

@@

Our main objective is to compute the cumulative incidence of relapse for a given set of covariates. We start by fitting a smooth hazard to the data using a linear term for time:

@@

We will compare our estimates to that obtained from a corresponding Fine-Gray model [-@fine1999proportional]. The Fine-Gray model is a semiparametric model for the cause-specific *subdistribution hazard*, i.e. the function $f_k(t)$ such that
$$CI_k(t) =1 - \exp\left( - \int_0^t f_k(u) \textrm{d}u \right),$$
where $CI_k(t)$ is the cause-specific cumulative incidence. The Fine-Gray model allows to directly assess the effect of a covariate on the subdistribution hazard, as opposed to the hazard. For the computation, we will use the \pkg{timereg} package [@timereg]:

@@

# 1. Cox with only sps
coxSPS <- survival::coxph(Surv(time = d.time, event = death) ~ sps, 
                          data = train, x = TRUE)
abcoxSPS <- survival::survfit(coxSPS, newdata = test)
# 2. Cox with only aps
coxAPS <- survival::coxph(Surv(time = d.time, event = death) ~ aps, 
                          data = train, x = TRUE)
abcoxAPS <- survival::survfit(coxAPS, newdata = test)
# 3. Cox with everything but sps and aps
coxFull <- survival::coxph(Surv(time = d.time, event = death) ~ . - aps - sps, 
                           data = train, x = TRUE)
abcoxFull <- survival::survfit(coxFull, newdata = test)
# 4. Kaplan-Meier
KM <- survival::coxph(Surv(time = d.time, event = death) ~ 1, data = test)
abKM <- survival::survfit(KM)
abKMcompare <- data.frame(
  Time = abKM$time,
  KM = 1 - abKM$surv
)

@@

```{r baselines, echo=FALSE, eval=eval_cs3}
cox_ci <- ggplot(baseCoxComparisonData, aes(Time)) +
  geom_line(aes(y = APS, colour = "APS"), lwd = 1) +
  geom_line(aes(y = Full, col = "Full"), lwd = 1) +
  geom_line(aes(y = SPS, colour = "SPS"), lwd = 1) +
  geom_step(data = abKMcompare, mapping = aes(x = Time, y = KM, color = "K-M"), lwd = 1) +
  labs(y = "Probability of Death", x = "Survival time (Years)", color = "Models") +
  scale_color_manual(values = q4, breaks = c("APS", "Full", "SPS", "K-M")) + 
  paper_gg_theme
```
